Bayesian regression modelling facilitates quantitative modelling of cell metabolism

### 2023-10-30 paper/main.qmd
Add acknowledgements, edit supplementary

---
title: "Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism 
abstract: >
  This paper presents Maud, a command-line application that implements
  Bayesian statistical inference for kinetic models of biochemical metabolic
  reaction networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
  review the existing options in this area, explain how Maud improves on the state
  of the art, describe the intended modelling workflow and illustrate its use with
  an example application.
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
    - id: biosustain
      name: DTU Biosustain
  - name: "Nicholas Luke Cowie"
    affiliations: 
    - ref: biosustain
  - name: "Lars Keld Nielsen"
    affiliations: 
    - ref: biosustain
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
      abbreviations: IR,NMR,UV
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

After specifying the qualitative aspects of the kinetic model, we selected
parameter values to use as ground truth by Monte Carlo sampling using a previous
model of the methionine cycle as a starting point (see @saa_general_2015 for
this model).

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_general_2015 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model described below in @sec-methionine-measurement-model.

@@

Running standard diagnostic checks indicated that the samples we generated
were from the target posterior distribution. The improved $\hat{R}$ statistic
[@vehtariRankNormalizationFoldingLocalization2021] for every sampled variable
was within 1% of 1, indicating appropriate mixing within and between Markov
chains. Additionally, the number of effective samples was high, indicating
that we generated enough posterior samples to support inferences about the
bulks of the distributions of the sampled parameters. Furthermore, we observed
no post warm-up divergent transitions, indicating that the sampler was able to
transform the log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

@@

>>>>

---
title: "Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism 
abstract: >
  This paper presents Maud, a command-line application that implements
  Bayesian statistical inference for kinetic models of biochemical metabolic
  reaction networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
  review the existing options in this area, explain how Maud improves on the state
  of the art, describe the intended modelling workflow and illustrate its use with
  an example application.
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
    - id: biosustain
      name: DTU
      department: DTU Biosustain
      city: Kongens Lyngby
      country: Denmark
  - name: "Nicholas Luke Cowie"
    affiliations: 
    - ref: biosustain
  - name: "Lars Keld Nielsen"
    affiliations: 
    - ref: biosustain
    - id: UQ
    - name: The University of Queensland
    - department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
    - city: St Lucia
    - country: Australia
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
      abbreviations: IR,NMR,UV
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

After specifying the qualitative aspects of the kinetic model, we selected
parameter values to use as ground truth by Monte Carlo sampling using a previous
model of the methionine cycle as a starting point (see @saa_construction_2016 for
this model).

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_construction_2016 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model described below in @sec-methionine-measurement-model.

@@

Running standard diagnostic checks indicated that the samples we generated
were from the target posterior distribution. The improved $\hat{R}$ statistic
[@vehtariRankNormalizationFoldingLocalization2021] for every variable of
interest was within 2% of 1, indicating appropriate mixing within and between
Markov chains. Additionally, the number of effective samples was high,
indicating that we generated enough posterior samples to support inferences
about the bulks of the distributions of the sampled parameters. Furthermore,
we observed no post warm-up divergent transitions, indicating that the sampler
was able to transform the log-posterior distribution, avoiding any regions with
excessive curvature that might inhibit exploration via HMC.

@@

# Supporting information

The document `supplementary.pdf` contains the following information:

- Description of Maud's kinetic model including parameter dimensions and rate
equations.
- Description of the procedure used to generate the artificial case study
datasets.
- Specification of the prior distribution used for the case studies

@@

This research was funded by the Novo Nordisk foundation using grant number
NNF14OC0009473.
